---
title: "Climate Change - Time Series Analysis(Global)"
subtitle: "DV Project"
output: html_document
---

# Faraz Suhail (19BCE1525), Hardik Govil (19BCE1530)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
par(mar=c(1,1,1,1))
library(tidyverse)
library(tseries)
library(fpp2)
library(gridExtra)
library(plotly)
```


### Structure of data
```{r, echo=FALSE}
rm(list=ls())
data = read.csv('GlobalTemperatures.csv')

str(data) # contains NA 
```

### Summary of data
```{r, echo=FALSE}
summary(data)
```

### Identify these NA values [Data Cleaning]
```{r, echo=TRUE}
#######################################################
# For Land Average Temperature and Uncertainty Column
#######################################################
sum(is.na(data[1:121,]$LandAverageTemperature))
sum(is.na(data[121:3192,]$LandAverageTemperature))
data$dt[121]
tail(data$dt,1)

#######################################################
# For other columns apart from Land Average Temp
#######################################################
sum(is.na(data$LandAndOceanAverageTemperature))
sum(is.na(data[1201:3192,]$LandMaxTemperature))
sum(is.na(data[1201:3192,]$LandMinTemperature))
sum(is.na(data[1201:3192,]$LandAndOceanAverageTemperature))
data$dt[1201]
tail(data$dt,1)
```

### Visualize the Time Series Data
```{r,echo=FALSE}
landAvgTemps <- data[121:3192,]$LandAverageTemperature
Y = ts(landAvgTemps,start=c(1760,1),frequency = 12)
print(paste("Start time:",start(Y)[1],start(Y)[2]))
print(paste("End time:",end(Y)[1],end(Y)[2]))

autoplot(Y , main="Time Series Plot [1760-2015]")
autoplot(decompose(Y))
```

##### INFERENCE: 
* A weak trend can be seen in decomposition plot.

#### Slicing the time period to see the curves properly
```{r,echo=FALSE}
######################################################
## Slicing the time period to see the curves properly
######################################################

autoplot(window(Y,1760,1800),main = "Time Series Plot [1760-1800]") + ylab("Temperature (C)")
#autoplot(window(Y,1800,1850),main = "Time Series Plot [1800-1850]")
autoplot(window(Y,1850,1900),main = "Time Series Plot [1850-1900]") + ylab("Temperature (C)")
#autoplot(window(Y,1900,1950),main = "Time Series Plot [1900-1950]") + ylab("Temperature (C)")
#autoplot(window(Y,1915,1965),main = "Time Series Plot [1915-1965]") + ylab("Temperature (C)")
autoplot(window(Y,1965,2015),main = "Time Series Plot [1965-2015]") + ylab("Temperature (C)")

```

##### INFERENCE: 
* Y-scale range changes according to the time period. 
* Some values are negative during 1760-1800.
* From 1850-1900 we can see all values are positive i.e. above 0°C.
* During 1965- 2015 the lowest values are found around 2°C.

```{r,echo=FALSE}
data %>%
  separate(col = dt, into = c("Year", "Month", "Day"), convert = TRUE) -> data_monthly 

data_monthly$Day=NULL
print("[INFO] Converted Date Column into Year and Month column for doing monthly and yearly analysis on the Time Series")

```


### Season wise comparison trend Visualization
```{r,echo=FALSE}

data_monthly = data_monthly %>%
  mutate(season=ifelse(Month %in% c(6,7,8),"Summer",
                       ifelse(Month %in% c(9,10,11),"Fall",
                              ifelse(Month %in% c(12,1,2),"Winter","Spring")
                       )))

ggplot(data = data_monthly, 
       aes(Year,LandAverageTemperature, colour=season)) +
  geom_point(na.rm = T) +
  xlab("Year") +
  ggtitle("Land Average temperature by year")
```

##### INFERENCE: 
* There is a distinct separation between Summer and Winter in terms of temperature range.
* Whereas Spring and Fall temperatures are almost in the same range.


### Month wise trend visualization in Average Land Temperature
```{r,echo=FALSE}

monthly_info = data_monthly %>%
  group_by(Month) %>%
  summarise(mean_temp = mean(LandAverageTemperature,na.rm=T),
            max_temp = max(LandAverageTemperature,na.rm=T),
            min_temp = min(LandAverageTemperature,na.rm=T))

monthly_info$Month_name = factor(1:12,labels=month.abb)

g = ggplot(monthly_info, aes(x=Month_name)) +
  geom_point(aes(y = mean_temp,color="mean_temp")) +
  geom_line(aes(y = mean_temp, group=1, color="mean_temp")) + geom_point(aes(y = max_temp, color = "max_temp")) +
  geom_line(aes(y = max_temp , group=1,  color="max_temp")) +
  geom_point(aes(y = min_temp, color = "min_temp")) +
  geom_line(aes(y = min_temp , group=1, color="min_temp") ) + scale_colour_manual("", breaks = c("mean_temp", "max_temp", "min_temp"),values = c("mean_temp"="forestgreen", "max_temp"="darkred", "min_temp"="steelblue")) +
  xlab("") + ylab("Temperature (C)") +
  labs(title = "Monthly Avg Global Land Temperature [Mean-Max-Min]")

ggplotly(g)
  

```


##### INFERENCE: 
* The temperature rises drastically from April and drops during October, this shows the seasonal nature of temperature throughout the year. 
* The hottest days occur during the month of July and coldest during January.


### Yearly trend Visualization
```{r,echo=FALSE}
data_yearly = data_monthly %>% 
  group_by(Year) %>% 
  summarise(LandAverageTemperature = mean(LandAverageTemperature,na.rm=T),
            LandAverageTemperatureUncertainty = mean(LandAverageTemperatureUncertainty,na.rm=T),
            LandMaxTemperature = mean(LandMaxTemperature,na.rm=T),
            LandMaxTemperatureUncertainty = mean(LandMaxTemperatureUncertainty,na.rm=T),
            LandMinTemperature = mean(LandMinTemperature,na.rm=T),
            LandMinTemperatureUncertainty = mean(LandMinTemperatureUncertainty,na.rm=T),
            LandAndOceanAverageTemperature = mean(LandAndOceanAverageTemperature,na.rm=T),
            LandAndOceanAverageTemperatureUncertainty = mean(LandAndOceanAverageTemperatureUncertainty,na.rm=T))


ggplot(data_yearly[101:266,], aes(x=Year, y=LandAverageTemperature)) +
  geom_point() + geom_smooth(method = 'loess',formula = 'y~x') +
  ggtitle("Yearly Global Land Average Temperature 1850-2015") +
  geom_errorbar(aes(x = Year,ymin = LandAverageTemperature - LandAverageTemperatureUncertainty,
                    ymax = LandAverageTemperature + LandAverageTemperatureUncertainty))+
  ylab("Average Temperature, Celsius")

# LandAverageTemperature
ggplot(data_yearly, aes(x=Year, y=LandAverageTemperature)) +
  geom_line() + geom_smooth(method = 'loess',formula = 'y~x') +
  ggtitle("Yearly Global Land Average Temperature 1750-2015")

# LandAndOCeanAverageTemperature
ggplot(data_yearly[101:266,], aes(x=Year, y=LandAndOceanAverageTemperature)) +
  geom_line() + geom_smooth(method = 'loess',formula = 'y~x') +
  ggtitle("Yearly Global Land And Ocean AverageTemperature 1850-2015")

```

##### INFERENCE: 
* All the graphs above shows subtle increase in the mean land temperature over the years. 
* This supports the fact of climate change that has been reported by many environmentalist.

### Comparison Max and Min Land Temperature
```{r,echo=FALSE}

# LandMaxTemperature
ggplot(data_yearly[101:266,], aes(x=Year, y=LandMaxTemperature)) +
  geom_line() + geom_smooth(method = 'loess',formula = 'y~x') +
  ggtitle("Yearly Global Land Max Temperature 1850-2015")

# LandMinTemperature
ggplot(data_yearly[101:266,], aes(x=Year, y=LandMinTemperature)) +
  geom_line() + geom_smooth(method = 'loess',formula = 'y~x') +
  ggtitle("Yearly Global Land Min Temperature 1850-2015")


q1 = qplot(Year, LandMaxTemperature, data=na.omit(data_yearly), main="Global Yearly Max Temperature 1850-2015",geom=c("point","smooth"))+ 
  aes(colour = LandMaxTemperature) + 
  scale_color_gradient(low="blue", high="red")

q2 = qplot(Year, LandMinTemperature, data=na.omit(data_yearly), main="Global Yearly Min Temperature 1850-2015",geom=c("point","smooth"))+ 
  aes(colour = LandMinTemperature) + 
  scale_color_gradient(low="cyan", high="blue")

grid.arrange(q1, q2,nrow=2, ncol=1)
```

##### INFERENCE:
* Temperatures have increased by a difference of approx 2°C in terms of both max and min temperature.
* Again this shows that there is an anomaly in mean temperature that corresponds to climate change.

### Trend And Seasonality investigation
```{r,echo=FALSE}
autoplot(window(Y,1990,2015)) + geom_smooth(method = 'loess', formula = 'y ~ x') + ggtitle("Time plot 1990-2015") + ylab("Avg Land Temperature (C)")

print("INFERENCE: Visually inspecting we dont see any trend in this Monthly Time Series")

ggseasonplot(window(Y,1990,2015)) + ggtitle("Seasonal plot [1990-2015]")

g1 = ggsubseriesplot(window(Y,1990,2015)) + ggtitle("Sub Series Plot [1990-2015]") + ylab("Avg Land Temperature (C)")

ggplotly(g1)

#plot(window(Y,1990,2015))
#title(main = "TS 1990-2015")
#abline(reg = lm(Y~time(Y)))

boxplot(window(Y,1990,2015)~cycle(window(Y,1990,2015)), ylab = "Avg Land Temperature (C)", xlab="Months", main="boxplot of Avg Land Temperature on Monthly basis")

print("Above three graphs: Seasonality plot, Subseriesplot and boxplot visualization shows whether the time series contains the seasonality component.")

interested_years = c(1860,1880,1900,1920,1940,1960,1980,2000,2015)
data_s = data_monthly %>%
  filter(Year %in% interested_years) %>%
           group_by(Year)

data_s$Year <- as.factor(data_s$Year)
qplot(x =  Year, y = LandAverageTemperature, data = data_s) +
  ggtitle("Average Temperature for 20 Year Intervals")+geom_boxplot(fill="turquoise")


# yearly trend
plot(aggregate(Y, FUN=mean),ylab="Temperature (C)")
title(main="Yearly trend plot")

plot(diff(aggregate(Y, FUN=mean)),ylab="Temperature (C)")
title(main="First Difference Yearly trend plot")

```

##### INFERENCE:
* There is no strong trend in Monthly Time Series which can be seen in the Time Plot 1990-2015.
* Strong Seasonality is present in the Monthly Time Series which is evident form seasonality plot.
* In the Box plot of "Average Temperature for 20 years" we can see gradual increase in the median temperatures over the 20 year interval.
* The yearly time series shows a slight positive trend.
* The first difference of yearly time series removes this trend.

### Apply forecasting ARIMA model on the ts data
* Two Direction: Monthly and Yearly forecasting

```{r,echo=FALSE}

monthly_ts = data_monthly$LandAverageTemperature[121:3192]
yearly_ts = data_yearly$LandAndOceanAverageTemperature[101:266]

Y_m = ts(monthly_ts,start=c(1760,1),frequency = 12)
#Y_y = ts(yearly_ts,start=c(1760),frequency = 1)
Y_y = ts(yearly_ts,start=c(1850),frequency = 1)

acf(Y_m, main="Monthly Time Series 1760-2015 [ACF]")

pacf(Y_m, main="Monthly Time Series 1760-2015 [PACF]")

acf(diff(Y_y), main="Yearly Time Series 1850-2015 [ACF]")

pacf(diff(Y_y), main="Yearly Time Series 1850-2015 [PACF]")

``` 


##### OBSERVATIONS:
* AR - p - PACF and MA - q - ACF
* Auto correlation is seen less in yearly Time series as most of lags are within the blue line

### ADF Test
```{r,echo=FALSE}

print("Monthly Time series")
adf.test(Y_m,alternative=c("stationary"))

print("Yearly Time series")
adf.test(Y_y,alternative = c("stationary"))

```

##### OBSERVATION:
* Monthly Time Series ~ p-value < 0.05 ~ alternate hypothesis accepted ~ stationary series 
* Yearly Time Series ~ p-value > 0.05 ~ cannot reject null hypothesis ~ non stationary series 

### BenchMark: SNAIVE method
* y_t = y_{t-s} + e_t

* This is a naive method in which we say the forecast value of this month will be the value of same month in previous year plus some error term.
```{r,echo=FALSE}

ts_1915_2015 = window(Y_m,1915,c(2015,12))
fit = snaive(ts_1915_2015)
print(paste("Residual SD:",sqrt(fit$model$sigma2)))
checkresiduals(fit)

```


#### INFERENCE:

* SNaive method was fitted to time series from 1915-2015 and we can see from residual plots that most of lags are outside blue lines hence the model is not able to fit the data properly.

* This gives us an idea of the benchmark performance for the forecasting model that we are going to try next.

* The p-value from Ljung-box test comes out to be very small hence autocorrelation is high in residuals which is evident from the ACF graph.


### ARIMA MODEL
```{r,echo=FALSE}

ts_1915_2015 = window(Y_m,1915,c(2015,12))
fit_arima_m = auto.arima(ts_1915_2015,D=1,trace =T,allowdrift = F)

print("Model Summary")
print(summary(fit_arima_m)) 
print(paste("Residual SD:",sqrt(fit_arima_m$sigma2)))
checkresiduals(fit_arima_m)

```
##### OBSERVATION:
* ARIMA model does a better job than Seasonal Naive method becuase the Residual SD of ARIMA is lower.

* The residual of ARIMA fitted good but still has some auto-correlation which can be seen in ACF as some lags are outside the blue line.

* Also Ljung-Box Test shows that p-value (< 0.05) is significant hence alternate hypothesis is accepted that means auto-correlation is present in the residuals.

* This tells us that there is an even better model can be made to that fits time series perfectly but it might be even more complicated than ARIMA.

```{r,echo=FALSE}
 
fcst = forecast(fit_arima_m,h = 60)
autoplot(fcst,include = 192,ylab="Average Land Temperature (C)", xlab="Year")

#print(paste("Mean Global Land Temperature 2020:",round(mean(tail(fcst$mean,12)),1)))

```


##### Monthly Forecast values for Year 2020
```{r,echo=F}
tail(fcst$mean,12)
```


#### ARIMA Model on Yearly TS
```{r,echo=FALSE}

fit_arima_y = auto.arima(Y_y,d=1,trace=T,allowdrift = T)

print("Model Summary")
print(summary(fit_arima_y))
print(paste("Residual SD:",sqrt(fit_arima_y$sigma2)))
checkresiduals(fit_arima_y)

```

##### OBSERVATION:

* The Ljung-Box Test shows that p-value (> 0.05) is not significant hence null hypothesis cannot be rejected  that means there isn't any strong auto-correlation in the residuals.

* This tells us the model fits the residual properly which is evident from ACF as almost all the lags are within the blue lines.

```{r,echo=FALSE}

fcst = forecast(fit_arima_y,h = 10,level=95)
autoplot(fcst, include=100, ylab="Average Land Ocean Temperature (C)", xlab="Year")

```


##### Yearly Forecast values till 2021
```{r,echo=FALSE}
fcst$mean

print(paste("Mean Global Land and Ocean Temperature 2020:",round(fcst$mean[5],4),"°C"))
print(paste("Mean Global Land and Ocean Temperature 2021:",round(fcst$mean[6],3),"°C"))
```

#### Conclusion:

* The Forecasted values in plot captures the trend and seasonality of the original time series.

* The Forecasted value for Mean Global Land and Ocean Temperature for year 2020 is '15.9532 °C' with an uncertainty of plus and minus of 1°C.

* The Forecasted value for 2021 is '15.9593 °C' which shows an increase in temperature from that of 2020

```{r,echo = T}
require(fGarch)
require(rugarch)
# okay to use the provided data - no transformation is needed
plot.ts(Y_y)
```

```{r,echo = T}
model2<-garchFit(~garch(1,1), data=Y_y, trace=F)
summary(model2)
predict(model2, n.ahead = 10, trace = FALSE)
```

```{r,echo = T}
model3<-garchFit(formula = ~arma(3,1,1)+garch(1,1), data=Y_y, trace=F)
summary(model3)
```


```{r}
m3 <-  ugarchspec(variance.model = list(model="fGARCH", 
                                        submodel = "GARCH", 
                                            garchOrder=c(1, 1)), 
                      mean.model = list(armaOrder=c(3,1,1)))
```

```{r}
fit <- ugarchfit(m3, Y_y,
                 solver = "hybrid")
print(fit)
fore <- ugarchforecast(fit, n.ahead = 10)
pred <- fore@forecast$seriesFor
pred
```

### FB's Prophet Model

* Prophet is a procedure for forecasting time series data based on an additive model where non-linear trends are fit with yearly, weekly, and daily seasonality, plus holiday effects.

* It works best with time series that have strong seasonal effects and several seasons of historical data.


```{r,echo=FALSE}
library(prophet)

# column : ds,y

df_y_ts = data.frame(ds=data_yearly[101:266,]$Year,y=data_yearly[101:266,]$LandAndOceanAverageTemperature)

df_y_ts$ds = as.character(df_y_ts$ds)
df_y_ts$ds <- as.Date(df_y_ts$ds, format="%Y")

prop_fit <- prophet(df_y_ts)
Future1 = make_future_dataframe(prop_fit,periods = 10, freq = "year")

forecast1 = predict(prop_fit,Future1)
prop_fit$component.modes
tail(forecast1[c('ds','yhat')])

dyplot.prophet(prop_fit,forecast1)
prophet_plot_components(prop_fit,forecast1)

```

##### OBSERVATION:

* Prophet model gives a forecast which is almost similar to ARIMA model's forecast.

* The interactive graph plot can be used to see the forecast values and trend for the future.

* Plot component function showcases the trend and yearly additive component of the time series.
